---
title: "Class 09: Candy Mini Project"
author: "Steven Studdard (PID: A19123470)"
format: pdf
toc: TRUE
---

## Importing the Candy Data

```{r}
#Load the dataset
candy <- read.csv("candy-data.txt", row.names = 1)

head(candy)
```

## What is in the dataset? 

```{r}
#Nrow to see rows, sum to sum up the T/F
nrow(candy)
sum(candy$fruity)
```

>Q1. How many different candy types are in this dataset?

There are 85 different candy types in this dataset. 

>Q2. How many fruity candy types are in the dataset?

There are 38 different fruity candy types in the dataset. 

## What is your favorite candy?


```{r}
#Load the dplyr library
library(dplyr)
```

```{r}
#We filter it to a singular row with the filter, and then select a specific area with select. We combine it with the pipes. 

candy|> 
  filter(row.names(candy)=="Twix") |> 
  select(winpercent)

candy|> 
  filter(row.names(candy)=="Snickers") |> 
  select(winpercent)

candy|> 
  filter(row.names(candy)=="Kit Kat") |> 
  select(winpercent)

candy|> 
  filter(row.names(candy)=="Tootsie Roll Snack Bars") |> 
  select(winpercent)
```

>Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

My favorite candy is Snickers and it's winpercent value is 76.67378. 

>Q4. What is the winpercent value for “Kit Kat”?

The winpercent value for Kit Kat is 76.7686. 

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

The winpercent value for Tootsie Roll Snack Bars is 49.6535. 

```{r}
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Winpercent is the only one that isn't on a 0-1 one scale. 

>Q7. What do you think a zero and one represent for the candy$chocolate column?

One means that it is true that the candy does contain chocolate, a zero means it does not. 

## Exploratory Analysis

```{r}
#Load ggplot
library(ggplot2)
```

>Q8. Plot a histogram of winpercent values using both base R and ggplot2.

```{r}
#Default R histogram
hist(candy$winpercent)

#ggplot histogram
ggplot(candy)+
  aes(winpercent)+
  geom_histogram()
```


>Q9. Is the distribution of winpercent values symmetrical?

No, the distribution of winpercent values are not symmetrical. 

>Q10. Is the center of the distribution above or below 50%?

It is below 50%. 

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

On average chocolate is higher ranked than fruit candy. 

```{r}
fruitywin <- mean(candy$winpercent[as.logical(candy$fruity)])
chocowin <- mean(candy$winpercent[as.logical(candy$chocolate)])

fruitywin > chocowin
```


>Q12. Is this difference statistically significant?

Yes, this difference is statistically significant. 

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)], candy$winpercent[as.logical(candy$fruity)])
```

## Overall Candy Rankings

```{r}
head(candy[order(candy$winpercent, decreasing = TRUE),], n=5)

head(candy[order(candy$winpercent),], n=5)

```

>Q13. What are the five least liked candy types in this set?

The 5 least liked candy types in this data set are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters. 

>Q14. What are the top 5 all time favorite candy types out of this set?

The top 5 all time candy types are Reese's Peanut Butter Cups, Reese's Miniatures, Twix, Kit Kat and Snickers. 

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r, fig.height = 15, fig.width = 10}
ggplot(candy)+
  aes(winpercent, row.names(candy))+
  geom_col()
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r, fig.height = 15, fig.width = 10}
ggplot(candy)+
  aes(winpercent, reorder(row.names(candy), winpercent))+
  geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r, fig.height = 15, fig.width = 10}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) +
  ylab("")
```

>Q17. What is the worst ranked chocolate candy?

Sixlets is the worst ranked chocolate candy. 

>Q18. What is the best ranked fruity candy?

Starburst is the best ranked fruity candy. 

## Taking a look at pricepercent

```{r}
library(ggrepel)
```

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures seem to be the most bang for your buck looking at the graph above. 

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

The top 5 most expensive candy types are Nik L Nip, Nestle Smarties, Ring Pop, Hershey's Krackel, Hershey's Milk Chocolate. The least popular of these is Nik L Nip. 

```{r}
candy |>
  arrange(-pricepercent) |> 
  select(pricepercent, winpercent) |> 
  head(n=5)
```

## Exploring the correlation structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

From the corrplot, it seems the two variables that are the most anti-correlated are chocolate and fruity.

>Q23. Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?

## Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
head(my_data)
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
```

```{r}
#ggplotly(p)
```

```{r}
ggplot(pca$rotation) +
  aes(PC1, reorder(row.names(pca$rotation), PC1)) +
  geom_col()
```

>Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

The three variables picked up strongly in the positive direction are fruity, pluribus, and hard. This does make sense as the three were slightly unique in the way that they had a negative correlation with most everything else, but positive with each other. 

## Summary

>Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

A winning candy is a chocolate bar with peanuts with an okay price point. Exploratory analysis helps with identifying that chocolate was statistically significant. Correlation analysis shows which traits cluster together. Then PCA gives us the structure and confirms what is driving the variance. It helps explain the previous two. 





















