---
title: "Class 5: Data Viz with ggplot2"
author: "Steven Studdard (PID: A19123470)"
format: pdf
toc: true
---

## Background

There are many graphics systems in R for making plots and figures. These include so-called *"base R" graphics* like the `plot()` function and add on packages like **ggplot2**. 

Let's compare how we make a simple figure with these two systems: 

We can use the in-built `cars` dataset:

```{r}
head(cars)
```
```{r}
plot(cars)
```

Before I can use ggplot2 I needto install it on my computer. 
To do this we can use the function `install.packages("ggplot2")`

> **N.B.** We never run `install.packages()` in our quarto doc (we run it once only in our console) as it would re-install the package every time we render our quarto report. 

```{r}
library(ggplot2)
```


The Main function in the **ggplot** package is called `ggplot()`.

Every ggplot has at least 3 layers

-the **data** (a data frame if the stuff we want to plot))
-the **aes**thetics (how the data maps to the plot)
-the "**geom** layer" (how you want the plot drawn, e.g points, lines, etc) 

```{r}

ggplot(data = cars)+
  aes(speed, dist)+
  geom_point()

```

## Add some custom features 

Let's add a trend line that shows the relationship between speed and distance. 

```{r}

ggplot(data = cars)+
  aes(x = speed, y = dist)+
  geom_point()+
  geom_smooth(se = FALSE, method = lm, col = 'yellow')+
  theme_minimal()+
  labs(title="Stopping Dist of Old Cars", 
       x = "Speed (MPH)", 
       y = "Distance (ft)")

```


Q. Can you make the `geom_smooth()` function produce a linear straight line fit to teh data and t urn off the "gray" error region.

---

## Import a Gene 


```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes, 5)
```

```{r}
sum(genes$State == "up")

```
A useful new function in this context is the `table()` function:

```{r}
table(genes$State)
```

First plot attempt

```{r}
ggplot(genes)+
  aes(Condition1, Condition2, col = State)+
  scale_color_manual(values = c("turquoise", "gray", "violet"))+
  geom_point()+
  theme_bw()+
  labs(x="No drug", y="Drug",
       title="Expression cahnges upon GLP-1 inhibitor treatment")
```

## Going Further

Here we read the famous gapmider dataset:

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

>Q. How many entries (i.e rows) are in this dataset?

```{r}
nrow(gapminder)
```

>Q. How many different "country" entries are in this data set. 

```{r}
length(table(gapminder$country))
```
```{r}
length(unique(gapminder$country))
```

Let's make our first plot of the entire dataset:

Plot of "gdpPercap" vs "ifeExp" colored by "continent" 

```{r}
p <- ggplot(gapminder)+
  aes(gdpPercap, lifeExp, col = continent)+
  geom_point(alpha=0.3)

p
```

I can add more layers to `p`
```{r}
p +
  facet_wrap(~gapminder$continent)
```

Make a plot for 1977 and 2007 only (not all the years in the dataset). 

First use the **dplyr** package and the `filter()` function from that package to extract the year 2007.

```{r, message=FALSE, warning=FALSE}
library(dplyr)
```

```{r}
g07 <- filter(gapminder, year == 2007)
g77 <- filter(gapminder, year == 1977)
g <- filter(gapminder, year == 2007 | year == 1997)
```

```{r}
ggplot(g)+
  aes(gdpPercap, lifeExp, col=continent, size=pop)+
  geom_point()+
  facet_wrap(~year)
```
>Q. Make a histogram of lifeExp colored by continent (try using `fill=continent` or `col=continent`)
>Q. Make a histogram of lifeExp facted by continent

```{r}
ggplot(gapminder)+
  aes(lifeExp, fill=continent)+
  geom_histogram()
```
```{r}
ggplot(gapminder)+
  aes(lifeExp, col=continent)+
  facet_wrap(~continent)+
  geom_histogram()
```

























