---
title: "Class:7 Machine Learning 1"
author: "Steven Studdard (PID: A19123470)"
format: pdf
toc: true
---

## Background

Today we will explore some core machine learning methods that are very popular in bioinformatics. These include **clustering** and **dimensionally reduction**.

## K-means clustering

The main function in "base" R for K-means clustering is called `kmeans

Before we get too deep let's make up some "simple" data that we cam cluster amd know if we are getting a good answer or not. To do this we an use the `rnorm()` function: 

```{r}
hist(rnorm(10000, mean = 3, sd = 2))
```

```{r}
#rnorm makes a random distribution 
#cbind takes sequence combines by columns
#rev reverses the argument
x <- c(rnorm(30,-3), rnorm(30,3))
z <- cbind(x=x,y=rev(x))
plot(z)
```

Now we can run `kmeans()` on tihs input `z` and see what the results look like. 

```{r}
km <- kmeans(z, centers = 2)
km
```

```{r}
attributes(km)
```

>Q How many points are in each cluster?

```{r}
km$size
```

>Q. What "componennt" of you result object details cluster assignemnet/membership?

```{r}
km$cluster
```

>Q. What "components of your result object details cluster center.

```{r}
km$centers
```

>Q. Plot `z` colored by the kmeanns cluster assignemnt and add cluster centers as blue points

```{r}
library(ggplot2)

plot(z, col=km$cluster)
points(km$centers, col="blue", pch=15)

#Annotate lets you do a simple correction to a geom
ggplot(z, aes(x, y, color =km$cluster))+
  geom_point()+
  annotate("point", km$centers[1,], km$centers[,-1], color = "red")+
  annotate("point", km$centers[-1,], km$centers[,1], color = "red")

```

>Q. Run a K-means clustering and plot the results asking for 4 clusters?

```{r}
k4m <- kmeans(z, centers=4)
k4m
plot(z, col=k4m$cluster, pch=16)
points(k4m$centers, col="blue", pch=15)

```

> **N.B.** You need to tell K-means the number of clusters (i.e. set `centers=2`)!!


One approach is to try different values for centers and then pick the best. 
```{r}

ans <- NULL
for(i in 1:10){
 km <- kmeans(z, centers=i)
 ans <- c(ans, km$tot.withinss)
}

plot(ans, type="o",
     xlab="Number of clusters",
     ylab="total Sum of Squares Distance")
```

## Hierarchical Clustering

The main function in "base" R for Hierarchical Clustering is called `hclust()`

The function does not take your "raw" data for clustering. You must first build a "distnace matrix" from your data and pass this as input to `hclust()`. 

```{r}
d <- dist(z)
hc <- hclust(d)
hc
```

There is a bespoke `plot()` method for `hclust()` result objects. 

```{r}
plot(hc)
abline(h=8, col="red")
```

Once we have our `hclust()` object (our "tree" of "cluster dendrogram") we can **cut** the tree to reveal the clustering pattern.

```{r}
cutree(hc, h=8)
cutree(hc, k=4)
```

>Q. Make a plot of `z` with your hclust results (i.e. color by cluster membership)

```{r}
grps <- cutree(hc, k=2)
plot(z, col= grps)
```

## Principal Component Analysis (PCA)

PCA is a dimensionality reduction method that is popular for revealing patterns in complex datasets. 

### Analysis of UK food data

Let's look at some data on the eating habits of folks from the UK to see if there are patterns and trends that have some regions being distinct from others. 

## Data Import

The data is made available in CSV format

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```

>Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

There are 17 rows and 5 columns we can find this out with nrow/ncol or just use dim(). 

```{r}
dim(x)
```

## Tidy the data

Fix anything that went wrong with the data import.

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
dim(x)
```

>Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I think that I prefer the method of just fixing the dataset as I import it in. When you re-run the code of subtracting one of the colunms to instead make it a row name, it slowly just removes more and more columns. You do not run into that issue if you just `x <- read.csv(url, row.names=1)` at the beginning. 

## Exporatory analysis

Make some plots to help make sense of obvious trends..

```{r}
barplot(as.matrix(x), beside=FALSE, col=rainbow(nrow(x)))
```
>Q3: Changing what optional argument in the above barplot() function results in the following plot?

Within the "base" R barplot function there is a argument that is called "beside" if beside is true then the columns of the data are shwon as seperate bars, but when it is false it is just stacked upon each other. 

```{r}
library(tidyr)
library(ggplot2)
# Convert data to long format for ggplot with `pivot_longer()`
x_long <- x |> 
          tibble::rownames_to_column("Food") |> 
          pivot_longer(cols = -Food, 
                       names_to = "Country", 
                       values_to = "Consumption")

dim(x_long)
head(x_long)

ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "stack") +
  theme_bw()
```
>Q4. Changing what optional argument in the above ggplot() code results in a stacked barplot figure?

You change the position argument in the `geom_col()`function. The position argument just changes how you want the bar plot to display your data. 

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

>Q5: We can use the pairs() function to generate ll pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

The `pairs()` function is creating a bunch of scatterplots with each dot being a different food category. Comparing each of our countries to one other country (for example the second plot is England compared with Wales). Being on the diagonal means that the two are very similar nearly 1:1 in what they eat. 

```{r}
library(pheatmap)

pheatmap( as.matrix(x) )
```
>Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

You can start to be able to tell the differences between the heatmap and the pairplots. N. Ireland seems to be seperate from the other three, most of the dots are off-center in the `pairs()` and in the tree its the longest branch. Its hard to be able to tell the main differences however, you have to look at it a bit.  

>**Key-point**: Even relatively small datasets can prove challenging to interpret. 

## PCA to the rescure

The main function in "base" R for PCA is caleld `prcomp()`. This function wants the "observations" to  be rows and the "variables" to be columns. 

So here we need to take the transpose of our `x` input object. 

```{r}
#T is transpose and it flips rows and col
pca <- prcomp(t(x))
summary(pca)
```

The returned `pca` object has components that we can use to make our main result figures. 

```{r}
attributes(pca)
```

The main result figure from this analysis is called a **"PC sore plot"** or "Ordination plot"or "PC Plot" or "PC1 vs PC2 plot".  

This plot shows how samples (in this case countries) relate to each other along our new PC axis. 

This is our new "reduced-dimensional space". In this case 2 dimensions, PC1 and PC2, that capture most of the variance in the original 17 dimension data-set. 

>Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
ggplot(pca$x)+
     aes(PC1, PC2)+
     geom_point()
```
>Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

```{r}
mycols <- c("orange", "red", "blue", "darkgreen")

ggplot(pca$x)+
     aes(PC1, PC2)+
     geom_point(col=mycols)
```
```{r}
ggplot(pca$x)+
     aes(PC1, PC2, label = row.names(pca$x))+
     geom_point(col=mycols)+
     geom_text(size = 3, vjust=2, col=mycols)
```

```{r}
ggplot(pca$rotation)+
  aes(PC1, reorder(row.names(pca$rotation), PC1))+
  geom_col()
```

>Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

```{r}
ggplot(pca$rotation)+
  aes(x=PC2, y= reorder(row.names(pca$rotation), PC2))+
  geom_col()
```

A PC score plot is a plot used to identify groupings in data. Creating "principal components" and mapping these sores out to see the variance in patterns. 

A loading plot is a plot to display the relationship between the original variables and hte principle components. 

A scree plot is used to analyze variance from the PC components of PCA. 



























